Should I AirBNB My Property?

### 2016-11-26 abbReturns_Script.Rmd
updating paths
filled out data prep

Data on AirBNB (short-term) rentals has been purchased from www.airdna.co, a data provider specializing in AirBNB data collection and analysis. The data on long-term leases was provided by Australia Property Monitors (APM), a Fairfax Group company.

Throughout this document we will refer to the short-term data as **'short-term'** or **'AirBNB'** while the long-term data will be referred to as **'long-term'**, **'rental'** or **'lease'**.  Short-term nightly fees will be referenced as **'rates'** and long-term as **'rent'**.  We will use **'prices'** to collectively refer to **rates** and **rents**. Note that within the Melbourne market rentals are quoted in weekly rents (but are charged to tenants at a monthly rent equivalent to the 1/12 of 52 weeks of rent).

@@

@@

Both the AirBNB and the rental data require some measure of data pre-processing and due to the run times of these function they are done outside of this document.  For the AirBNB data, the pre-processing is primarily concerned with imputing missing daily observations for a small number of properties in the dataset.  Full information on this process can be found in the **airbnbDataClean.Rmd** and **dataCleanFunctions.R** files.  Do note that most of the imputed data is ignored by our choice of study time period later on.

@@

  # Daily rental information

@@

We then load the long term rental data.  This includes property level data (including the final lease rate) and listing-level data containing information on all changes to listing prices and dates, etc.

```{r load_rent_data, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 ## Property level long term rental information

  rent.data <- read.csv(paste0(data.path, '/prepared/ltpropdata.csv'), header=T)

 ## Long term rental listing information

@@

  names(rent.data) <- tolower(names(rent.data))
  names(listing.data) <- tolower(names(listing.data))

@@

  rent.keep.flds <- c('id.key', 'geographicalid', 'eventid', 'addressid', 'activityid',
                      'flatnumber', 'streetnumber', 'streetname', 'streettype', 
                      'suburb', 'postcode', 'property_latitude',
                      'property_longitude', 'street_centroid_latitude',
                      'street_centroid_longitude', 'eventdate', 'eventprice',
                      'propertytype', 'areasize', 'bedrooms', 'baths', 'parking',
                      'hasstudy', 'hascourtyard', 'hasbalcony', 'hasairconditioning',
                      'hasgarage', 'dom')

  rent.data <- rent.data[ , rent.keep.flds]

@@

Periods are inserted between relevant words for the rental data.

@@

  names(rent.data) <- c('id.key', 'geographical.id', 
                        'event.id', 'address.id', 'activity.id',
                        'flat.number', 'street.number', 'street.name', 'street.type',
                        'suburb', 'postcode', 'latitude', 'longitude',
                        'street.latitude', 'street.longitude', 'event.date',
                        'event.price', 'property.type', 'area.size', 'bedrooms',
                        'baths', 'parking', 'has.study', 'has.courtyard',
                        'has.balcony', 'has.aircond', 'has.garage')

@@

  # Date
  rent.data$event.date <- fixLTDates(rent.data$event.date)

  # Year
  rent.data$event.year <- as.numeric(substr(rent.data$event.date, 1, 4))

@@

  rent.data$has.study <- ifelse(rent.data$has.study == 'True', 1, 0)
  rent.data$has.courtyard <- ifelse(rent.data$has.courtyard == 'True', 1, 0)
  rent.data$has.aircon <- ifelse(rent.data$has.aircon == 'True', 1, 0)
  rent.data$has.balcony <- ifelse(rent.data$has.balcony == 'True', 1, 0)
  rent.data$has.garage <- ifelse(rent.data$has.garage == 'True', 1, 0)

@@

AirBNB units are classified into 19 different property type.  The most common are Apartments and Houses, the least common are Igloo, Tent, Treehouse and Yurt^[Property types are owner/lister defined, therfore it is likely that the Igloo, at the very least, is not a completely true representation of the property type]. For the sake of this analysis, we will collapse these twenty types into three types:

@@

Within the long-term rental data, property types fall into eight categories.  Like the AirBNB data we group these into three categories matching as best as possible to the AirBNB categories:

@@

Part of the difficulty in perfectly mapping property types from these two datasets is due to the fact that the long-term data uses Australian terms while the AirBNB data conforms to North American lexicon.  This is most apparent when talking about apartment dwellings.  In Australia these are referred to as Units, while in North American they are referred to as Apartments or Condominiums (depending on ownership structure).  Additionally, Terrace homes are very common in many of the older suburbs of Melbourne and would be considered Rowhouses or, likely, Townhomes in the North American context.

@@

  rent.data$type <- 'Other'
  rent.data$type[rent.data$property.type %in% c('Duplex', 'House',
                                           'Terrace', 'Townhouse')] <- 'House'
  rent.data$type[rent.data$property.type %in% c('Studio', 'Unit')] <- 'Apartment'

@@

The **Other** categories contain very small percentage of properties in both the AirBNB and the rental data.  As a result, we remove the 'Others' from the data.

@@

  abb.data <- subset(abb.data, type != 'Other')
  rent.data <- subset(rent.data, type != 'Other')

@@

Next, for the short term rentals units can be listed as one of three types relating to the extent of the property which is able to be booked:

@@

As long-term rentals do not offer **Private Room** or **Shared Room** options and our purpose here is a comparison of the two we remove all short term properties that do not lease the entire home/apt. Unfortunately, this filter does remove about 40% of the short term data, however, given the research question it is unavoidable.

@@

  ggplot(abb.data, aes(x=listing.type)) + 
    geom_bar(color=abb.col[1], fill=abb.col[1]) +
    xlab('\nListing Type\n') + 
    ggtitle('\nShort Term Rentals\n')  +
    ylab('# of Properties\n')

@@

Next we trim the observations by time.  Our hypothetical simulated decision is that of an investor who purchases a property on September 1, 2015 and must decide on traditional long term rental versus an AirBNB, short term rental approach to generating income.  To determine the likely long term rental rate for any given property, we will use the observed rental transactions from the market from September 1, 2014 to August 31, 2015.  Taking the AirBNB approach allows the owner to change rates over time and is subjected to daily changes in the market (supply and demand factor).  In this simulated example we will use AirBnB data from September 1, 2015 to August 31, 2016 to represent the actual AirBNB market conditions over the time period in question.  Another reason for using these two period of data is that the AirBNB data has missing daily values prior to September 1, 2016.  While we have imputed these missing daily observations, we hold the actual observed values to be of greater accuracy and have designed the initial simulated example to take advantage of the full observed set.

@@

  rent.dataf <- rent.data[rent.data$event.date >= '2014-09-01' & 
                            rent.data$event.date <= '2015-08-31', ]
  listing.dataf <- listing.data[listing.data$id.key %in% rent.dataf$id.key, ]

@@

Location is clearly an important determinant in both the short and long term rental markets. At the broadest scale, the long-term Melbourne residential market is usually discussed in terms of Inner, Middle and Outer suburbs.  In general, prices are highest in the inner suburbs and lowest in the outer, with a number of exceptions in the high end neighborhoods in the east and southeastern areas of the middle suburbs.  Suburbs in Melbourne are much smaller than their North American counterparts.  The specific suburb of a property is the second, or finer scale, at which the market operates. The plot below shows the three large submarkets.

@@

  # Limit to extent of data
  sub.table <- table(rent.dataf$suburb)
  sub.table <- sub.table[sub.table != 0]

  # Clip to only those with a rental observation
  study.suburbs <- suburbs.shp[suburbs.shp@data$name_2006 %in% names(sub.table),]

@@

It is likely that the short-term rental market follows a similar market hierarchy, however, we can imagine that a few different spatial features can influence the short-term market.  While there are many possible additional features, we consider two to be the most likely to influence rates or occupancy: 1) Proximity to beach; and 2) Proximity to key tourist activities and events.  Using the three broad submarkets from the long-term market (inner, middle and outer) we have created a five sub-market system as a starting point for spatially analysing short and long term rentals in Melboure:

@@

About 5% of the long term rental observations have missing latitude and longitude values.  Half of these do, however, have a lat/long value for the centroid of the street that the property faces.  To retain as many observations as possible in these intial steps for those properties that lack a specific lat/long value we apply the street centroid value.  For those without either (3% of the total) we remove them from the dataset at this point.  After removing these, we apply the submarket designations.

@@

  miss.LL <- which(is.na(rent.dataf$latitude))

@@

  rent.dataf$miss.LL <- 0
  rent.dataf$miss.LL[miss.LL] <- 1

@@

  rent.dataf$latitude[miss.LL] <- rent.dataf$street.latitude[miss.LL]
  rent.dataf$longitude[miss.LL] <- rent.dataf$street.longitude[miss.LL]

@@

  rent.dataf <- rent.dataf[!is.na(rent.dataf$latitude), ]

@@

  # Convert to point shapefile
  rent.prop.shp <- SpatialPointsDataFrame(cbind(rent.dataf$longitude,
                                                rent.dataf$latitude),
                                         rent.dataf,
                                         proj4string=CRS("+init=epsg:4283"))

  # Add suburb name 
  spJoin <- over(rent.prop.shp, suburbs.shp)
  rent.dataf$suburb <- as.character(spJoin$name_2006)
  rent.dataf$suburb[is.na(rent.dataf$suburb)] <- 'missing'

  # Add suburb designation
  rent.dataf$sub.class <- as.character(spJoin$suburbdesi)
  rent.dataf$sub.class[is.na(rent.dataf$sub.class)] <- 'missing'

  # Add SA1 designation
  spJoin <- over(rent.prop.shp, sa1s.shp)
  rent.dataf$sa1 <- spJoin$sa1_main11

  ## Add Beach Designation
  spJoin <- over(rent.prop.shp, beach.buffer)
  rent.dataf$beach <- ifelse(is.na(spJoin), 0 ,1)

  # Remove those missing a suburb classification (out of area)
  rent.dataf <- rent.dataf[rent.dataf$sub.class != 'missing', ]
  rent.dataf <- rent.dataf[rent.dataf$sa1 != 'missing', ]

@@

  # Add individual classes
  rent.dataf$sub.mrkt <- 'rural'
  rent.dataf$sub.mrkt[rent.dataf$sub.class == 'Middle'] <- 'suburban'
  rent.dataf$sub.mrkt[rent.dataf$sub.class == 'Inner'] <- 'city'
  rent.dataf$sub.mrkt[rent.dataf$suburb %in% core.suburbs] <- 'city-core'
  rent.dataf$sub.mrkt[rent.dataf$beach == 1] <- 'beach'

  # Order the factors properly
  rent.dataf$sub.mrkt <- factor(rent.dataf$sub.mrkt, 
                               levels=c('city-core', 'city', 'suburban', 
                                        'rural', 'beach'))

@@

```{r rent.plot, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  # Make map
  rent.map <- ggplot() +
    geom_polygon(data=subs, aes(x=long, y=lat, group=group), color='gray50') +
    geom_point(data=rent.dataf, aes(x=longitude, y=latitude, color=sub.mrkt), size=.1) +
    scale_color_manual(values=sm.col) +
    scale_x_continuous(limits=c(min(subs$long), max(subs$long))) +
    scale_y_continuous(limits=c(min(subs$lat), max(subs$lat))) +
    xlab('') + ylab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank()) +
    guides(colour = guide_legend(override.aes = list(size=3))) + 
    ggtitle('Long Term Rental Location\nBy Submarket')

@@

  ggMultiPlots(abb.map, rent.map, cols=2)

@@

  # Make map
  renttype.map <- ggplot() +
    geom_bar(data=rent.dataf, aes(x=sub.mrkt, fill=sub.mrkt)) +
    scale_fill_manual(values=sm.col) +
    facet_wrap(~type) +
    ylab('# of Properties\n') + xlab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle('Long Term Property Type\nBy Submarket')

@@

  ggMultiPlots(abbtype.map, renttype.map, cols=2)

@@

We now filter both datasets by removing observed values unlikely to represent 'typical' short or long term rental properties.

@@

We start by looking at bedrooms. We begin by removing all properties that are missing bedrooms information or have a likely data error in this fields (more than 14 bedrooms). We then plot the distribution of different bedrooms counts.  From this point onwards we bifurcate our analysis between houses and apartments as the markets (prices, rates, and properties) differ between the two property types.

@@

  # Long term data
  rent.dataf <- rent.dataf[!is.na(rent.dataf$bedrooms), ]
  rent.dataf <- rent.dataf[rent.dataf$bedrooms <= 10, ]

@@

  rentbed <- ggplot(rent.dataf, aes(x=bedrooms)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:10, breaks=1:10)+
    xlab('\nBedrooms\n') +
    ylab('') +
    ggtitle('Bedrooms in Long Term Rentals')

@@

  abbbed <- ggplot(abb.dataf, aes(x=bedrooms)) + 
    geom_bar(fill=abb.col[1], color=abb.col[1]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=0:10, breaks=0:10)+
    xlab('\nBedrooms\n') +
    ylab('') +
    ggtitle('Bedrooms in Short Term Rentals')

@@

  ggMultiPlots(abbbed, rentbed, cols=1)

@@

For both short and long term rentals we see that most apartments have less than 4 bedrooms and houses less than 5.  We filter the data accordingly. Additionally, note that short term rentals consider studio apartments to be 0 bed units while the long term data considers these as 1 bedroom. We convert the 0 bed short term units to 1 bed units.

@@

  abb.dataf <- abb.dataf[abb.dataf$bedrooms < 5, ]
  rent.dataf <- rent.dataf[rent.dataf$bedrooms < 5, ]

@@

  abba <- which(abb.dataf$bedrooms > 3 & abb.dataf$type =='Apartment')
  renta <- which(rent.dataf$bedrooms > 3 & rent.dataf$type =='Apartment')
  if(length(abba)>0) abb.dataf <- abb.dataf[-abba, ]
  if(length(renta)>0) rent.dataf <- rent.dataf[-renta, ]

@@

 ## Remove missing data
rent.dataf <- rent.dataf[!is.na(rent.dataf$baths), ]
abb.dataf <- abb.dataf[!is.na(abb.dataf$baths), ]

@@

  ltbath <- ggplot(rent.dataf, aes(x=baths)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:8, breaks=1:8)+
    xlab('\nBathrooms\n') +
    ylab('') + 
    ggtitle('Bathrooms in Long Term Rentals')

@@

  stbath <- ggplot(abb.dataf, aes(x=baths)) + 
    geom_bar(fill=abb.col[1], color=abb.col[1]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=0:8, breaks=0:8)+
    xlab('\nBathrooms\n') +
    ylab('') + 
    ggtitle('Bathrooms in Short Term Rentals')

@@

  # Long term data
  rent.dataf <- rent.dataf[rent.dataf$baths < 4, ]

@@

In addition to looking at bedrooms and bathrooms as separate dimensions of properties, they can also be considered together.  In other words, it is very common in the industry (and when looking for a short term rental) to specify searches by bed/bath combination for obvious reasons -- 1 bedroom with 3 baths is inefficient and 4 bedrooms with 1 bath is uncomfortable.  As a result we examine the combinations of the two.

@@

  rent.dataf$bedbath <- paste0(rent.dataf$bedrooms, '..', rent.dataf$baths)
  abb.dataf$bedbath <- paste0(abb.dataf$bedrooms, '..', abb.dataf$baths)

@@

  ltb <- ggplot(rent.dataf, aes(x=bedbath)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    xlab('\nBed/Bath Combo\n') +
    ylab('Number of Properties\n') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle('Long Term Rentals') + 
    facet_wrap(~type)

@@

  stb <- ggplot(abb.dataf, aes(x=bedbath)) +
    geom_bar(fill=abb.col[1], color=abb.col[1]) + 
    xlab('\nBed/Bath Combo\n') +
    ylab('Number of Properties\n') +
    ggtitle('Short Term Rentals') + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    facet_wrap(~type)

@@

  rent.dataf <- rent.dataf[rent.dataf$bedbath %in% acc.bb, ]
  abb.dataf <- abb.dataf[abb.dataf$bedbath %in% acc.bb, ]

@@

### Rates and Rents

Finally, we turn to filtering by the prices -- the nightly rates and weekly rental values.  We start by removing observations missing price data.

@@

  rent.dataf <- rent.dataf[!is.na(rent.dataf$event.price), ]
  abb.dataf <- abb.dataf[!is.na(abb.dataf$nightly.rate), ]

@@

  # Rent Plot
  rent.plot <- ggplot(rent.dataf, aes(x=event.price)) + 
    geom_density(fill=abb.col[3],alpha=.8) +
    ylab('') +
    xlab('Weekly Rent') + 
    ggtitle('Long Term Rentals') + 
    scale_x_continuous(limits=c(0, 1500)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank())

  # Nightly Rate plot
  rate.plot <- ggplot(abb.dataf, aes(x=nightly.rate)) + 
    geom_density(fill=abb.col[1], alpha=.8) + 
    ylab('') +
    xlab('Nightly Rate') +
    ggtitle('Short Term Rentals') + 
    scale_x_continuous(limits=c(0, 600)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank())

>>>>

Data on AirBNB (short-term) ltrals has been purchased from www.airdna.co, a data provider specializing in AirBNB data collection and analysis. The data on long-term leases was provided by Australia Property Monitors (APM), a Fairfax Group company.

Throughout this document we will refer to the short-term data as **'short-term'** or **'AirBNB'** while the long-term data will be referred to as **'long-term'**, **'ltral'** or **'lease'**.  Short-term nightly fees will be referenced as **'rates'** and long-term as **'ltr'**.  We will use **'prices'** to collectively refer to **rates** and **ltrs**. Note that within the Melbourne market ltrals are quoted in weekly ltrs (but are charged to tenants at a monthly ltr equivalent to the 1/12 of 52 weeks of ltr).

@@

 ## Set a global parameter

  exch.rate <- 1.32

@@

Both the AirBNB and the ltral data require some measure of data pre-processing and due to the run times of these function they are done outside of this document.  For the AirBNB data, the pre-processing is primarily concerned with imputing missing daily observations for a small number of properties in the dataset.  Full information on this process can be found in the **airbnbDataClean.Rmd** and **dataCleanFunctions.R** files.  Do note that most of the imputed data is ignored by our choice of study time period later on.

@@

  # Daily ltral information

@@

We then load the long term ltral data.  This includes property level data (including the final lease rate) and listing-level data containing information on all changes to listing prices and dates, etc.

```{r load_ltr_data, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 ## Property level long term ltral information

  ltr.data <- read.csv(paste0(data.path, '/prepared/ltpropdata.csv'), header=T)

 ## Long term ltral listing information

@@

  names(ltr.data) <- tolower(names(ltr.data))
  names(listing.data) <- tolower(names(listing.data))

@@

  ltr.keep.flds <- c('id.key', 'geographicalid', 'eventid', 'addressid', 'activityid',
                      'flatnumber', 'streetnumber', 'streetname', 'streettype', 
                      'suburb', 'postcode', 'property_latitude',
                      'property_longitude', 'street_centroid_latitude',
                      'street_centroid_longitude', 'eventdate', 'eventprice',
                      'propertytype', 'areasize', 'bedrooms', 'baths', 'parking',
                      'hasstudy', 'hascourtyard', 'hasbalcony', 'hasairconditioning',
                      'hasgarage', 'dom')

  ltr.data <- ltr.data[ , ltr.keep.flds]

@@

Periods are inserted between relevant words for the ltral data.

@@

  names(ltr.data) <- c('id.key', 'geographical.id', 
                        'event.id', 'address.id', 'activity.id',
                        'flat.number', 'street.number', 'street.name', 'street.type',
                        'suburb', 'postcode', 'latitude', 'longitude',
                        'street.latitude', 'street.longitude', 'event.date',
                        'event.price', 'property.type', 'area.size', 'bedrooms',
                        'baths', 'parking', 'has.study', 'has.courtyard',
                        'has.balcony', 'has.aircond', 'has.garage', 'dom')

@@

  # Date
  ltr.data$event.date <- fixLTDates(ltr.data$event.date)

  # Year
  ltr.data$event.year <- as.numeric(substr(ltr.data$event.date, 1, 4))

@@

  ltr.data$has.study <- ifelse(ltr.data$has.study == 'True', 1, 0)
  ltr.data$has.courtyard <- ifelse(ltr.data$has.courtyard == 'True', 1, 0)
  ltr.data$has.aircon <- ifelse(ltr.data$has.aircon == 'True', 1, 0)
  ltr.data$has.balcony <- ifelse(ltr.data$has.balcony == 'True', 1, 0)
  ltr.data$has.garage <- ifelse(ltr.data$has.garage == 'True', 1, 0)

@@

AirBNB units are classified into 19 diffeltr property type.  The most common are Apartments and Houses, the least common are Igloo, Tent, Treehouse and Yurt^[Property types are owner/lister defined, therfore it is likely that the Igloo, at the very least, is not a completely true representation of the property type]. For the sake of this analysis, we will collapse these twenty types into three types:

@@

Within the long-term ltral data, property types fall into eight categories.  Like the AirBNB data we group these into three categories matching as best as possible to the AirBNB categories:

@@

Part of the difficulty in perfectly mapping property types from these two datasets is due to the fact that the long-term data uses Australian terms while the AirBNB data conforms to North American lexicon.  This is most appaltr when talking about apartment dwellings.  In Australia these are referred to as Units, while in North American they are referred to as Apartments or Condominiums (depending on ownership structure).  Additionally, Terrace homes are very common in many of the older suburbs of Melbourne and would be considered Rowhouses or, likely, Townhomes in the North American context.

@@

  ltr.data$type <- 'Other'
  ltr.data$type[ltr.data$property.type %in% c('Duplex', 'House',
                                           'Terrace', 'Townhouse')] <- 'House'
  ltr.data$type[ltr.data$property.type %in% c('Studio', 'Unit')] <- 'Apartment'

@@

The **Other** categories contain very small percentage of properties in both the AirBNB and the ltral data.  As a result, we remove the 'Others' from the data.

@@

  abb.data <- subset(abb.data, type != 'Other')
  ltr.data <- subset(ltr.data, type != 'Other')

@@

Next, for the short term ltrals units can be listed as one of three types relating to the extent of the property which is able to be booked:

@@

As long-term ltrals do not offer **Private Room** or **Shared Room** options and our purpose here is a comparison of the two we remove all short term properties that do not lease the entire home/apt. Unfortunately, this filter does remove about 40% of the short term data, however, given the research question it is unavoidable.

@@

  ggplot(abb.data, aes(x=listing.type)) + 
    geom_bar(color=abb.col[1], fill=abb.col[1]) +
    xlab('\nListing Type\n') + 
    ggtitle('\nShort Term ltrals\n')  +
    ylab('# of Properties\n')

@@

Next we trim the observations by time.  Our hypothetical simulated decision is that of an investor who purchases a property on September 1, 2015 and must decide on traditional long term ltral versus an AirBNB, short term ltral approach to generating income.  To determine the likely long term ltral rate for any given property, we will use the observed ltral transactions from the market from September 1, 2014 to August 31, 2015.  Taking the AirBNB approach allows the owner to change rates over time and is subjected to daily changes in the market (supply and demand factor).  In this simulated example we will use AirBnB data from September 1, 2015 to August 31, 2016 to represent the actual AirBNB market conditions over the time period in question.  Another reason for using these two period of data is that the AirBNB data has missing daily values prior to September 1, 2016.  While we have imputed these missing daily observations, we hold the actual observed values to be of greater accuracy and have designed the initial simulated example to take advantage of the full observed set.

@@

  ltr.dataf <- ltr.data[ltr.data$event.date >= '2014-09-01' & 
                            ltr.data$event.date <= '2015-08-31', ]
  listing.dataf <- listing.data[listing.data$id.key %in% ltr.dataf$id.key, ]

@@

Location is clearly an important determinant in both the short and long term ltral markets. At the broadest scale, the long-term Melbourne residential market is usually discussed in terms of Inner, Middle and Outer suburbs.  In general, prices are highest in the inner suburbs and lowest in the outer, with a number of exceptions in the high end neighborhoods in the east and southeastern areas of the middle suburbs.  Suburbs in Melbourne are much smaller than their North American counterparts.  The specific suburb of a property is the second, or finer scale, at which the market operates. The plot below shows the three large submarkets.

@@

  # Limit to extent of data
  sub.table <- table(ltr.dataf$suburb)
  sub.table <- sub.table[sub.table != 0]

  # Clip to only those with a ltral observation
  study.suburbs <- suburbs.shp[suburbs.shp@data$name_2006 %in% names(sub.table),]

@@

It is likely that the short-term ltral market follows a similar market hierarchy, however, we can imagine that a few diffeltr spatial features can influence the short-term market.  While there are many possible additional features, we consider two to be the most likely to influence rates or occupancy: 1) Proximity to beach; and 2) Proximity to key tourist activities and events.  Using the three broad submarkets from the long-term market (inner, middle and outer) we have created a five sub-market system as a starting point for spatially analysing short and long term ltrals in Melboure:

@@

About 5% of the long term ltral observations have missing latitude and longitude values.  Half of these do, however, have a lat/long value for the centroid of the street that the property faces.  To retain as many observations as possible in these intial steps for those properties that lack a specific lat/long value we apply the street centroid value.  For those without either (3% of the total) we remove them from the dataset at this point.  After removing these, we apply the submarket designations.

@@

  miss.LL <- which(is.na(ltr.dataf$latitude))

@@

  ltr.dataf$miss.LL <- 0
  ltr.dataf$miss.LL[miss.LL] <- 1

@@

  ltr.dataf$latitude[miss.LL] <- ltr.dataf$street.latitude[miss.LL]
  ltr.dataf$longitude[miss.LL] <- ltr.dataf$street.longitude[miss.LL]

@@

  ltr.dataf <- ltr.dataf[!is.na(ltr.dataf$latitude), ]

@@

  # Convert to point shapefile
  ltr.prop.shp <- SpatialPointsDataFrame(cbind(ltr.dataf$longitude,
                                                ltr.dataf$latitude),
                                         ltr.dataf,
                                         proj4string=CRS("+init=epsg:4283"))

  # Add suburb name 
  spJoin <- over(ltr.prop.shp, suburbs.shp)
  ltr.dataf$suburb <- as.character(spJoin$name_2006)
  ltr.dataf$suburb[is.na(ltr.dataf$suburb)] <- 'missing'

  # Add suburb designation
  ltr.dataf$sub.class <- as.character(spJoin$suburbdesi)
  ltr.dataf$sub.class[is.na(ltr.dataf$sub.class)] <- 'missing'

  # Add SA1 designation
  spJoin <- over(ltr.prop.shp, sa1s.shp)
  ltr.dataf$sa1 <- spJoin$sa1_main11

  ## Add Beach Designation
  spJoin <- over(ltr.prop.shp, beach.buffer)
  ltr.dataf$beach <- ifelse(is.na(spJoin), 0 ,1)

  # Remove those missing a suburb classification (out of area)
  ltr.dataf <- ltr.dataf[ltr.dataf$sub.class != 'missing', ]
  ltr.dataf <- ltr.dataf[ltr.dataf$sa1 != 'missing', ]

@@

  # Add individual classes
  ltr.dataf$sub.mrkt <- 'rural'
  ltr.dataf$sub.mrkt[ltr.dataf$sub.class == 'Middle'] <- 'suburban'
  ltr.dataf$sub.mrkt[ltr.dataf$sub.class == 'Inner'] <- 'city'
  ltr.dataf$sub.mrkt[ltr.dataf$suburb %in% core.suburbs] <- 'city-core'
  ltr.dataf$sub.mrkt[ltr.dataf$beach == 1] <- 'beach'

  # Order the factors properly
  ltr.dataf$sub.mrkt <- factor(ltr.dataf$sub.mrkt, 
                               levels=c('city-core', 'city', 'suburban', 
                                        'rural', 'beach'))

@@

```{r ltr.plot, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  # Make map
  ltr.map <- ggplot() +
    geom_polygon(data=subs, aes(x=long, y=lat, group=group), color='gray50') +
    geom_point(data=ltr.dataf, aes(x=longitude, y=latitude, color=sub.mrkt), size=.1) +
    scale_color_manual(values=sm.col) +
    scale_x_continuous(limits=c(min(subs$long), max(subs$long))) +
    scale_y_continuous(limits=c(min(subs$lat), max(subs$lat))) +
    xlab('') + ylab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank()) +
    guides(colour = guide_legend(override.aes = list(size=3))) + 
    ggtitle('Long Term ltral Location\nBy Submarket')

@@

  ggMultiPlots(abb.map, ltr.map, cols=2)

@@

  # Make map
  ltrtype.map <- ggplot() +
    geom_bar(data=ltr.dataf, aes(x=sub.mrkt, fill=sub.mrkt)) +
    scale_fill_manual(values=sm.col) +
    facet_wrap(~type) +
    ylab('# of Properties\n') + xlab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle('Long Term Property Type\nBy Submarket')

@@

  ggMultiPlots(abbtype.map, ltrtype.map, cols=2)

@@

We now filter both datasets by removing observed values unlikely to represent 'typical' short or long term ltral properties.

@@

We start by looking at bedrooms. We begin by removing all properties that are missing bedrooms information or have a likely data error in this fields (more than 14 bedrooms). We then plot the distribution of diffeltr bedrooms counts.  From this point onwards we bifurcate our analysis between houses and apartments as the markets (prices, rates, and properties) differ between the two property types.

@@

  # Long term data
  ltr.dataf <- ltr.dataf[!is.na(ltr.dataf$bedrooms), ]
  ltr.dataf <- ltr.dataf[ltr.dataf$bedrooms <= 10, ]

@@

  ltrbed <- ggplot(ltr.dataf, aes(x=bedrooms)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:10, breaks=1:10)+
    xlab('\nBedrooms\n') +
    ylab('') +
    ggtitle('Bedrooms in Long Term ltrals')

@@

  abbbed <- ggplot(abb.dataf, aes(x=bedrooms)) + 
    geom_bar(fill=abb.col[1], color=abb.col[1]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=0:10, breaks=0:10)+
    xlab('\nBedrooms\n') +
    ylab('') +
    ggtitle('Bedrooms in Short Term ltrals')

@@

  ggMultiPlots(abbbed, ltrbed, cols=1)

@@

For both short and long term ltrals we see that most apartments have less than 4 bedrooms and houses less than 5.  We filter the data accordingly. Additionally, note that short term ltrals consider studio apartments to be 0 bed units while the long term data considers these as 1 bedroom. We convert the 0 bed short term units to 1 bed units.

@@

  abb.dataf <- abb.dataf[abb.dataf$bedrooms < 5, ]
  ltr.dataf <- ltr.dataf[ltr.dataf$bedrooms < 5, ]

@@

  abba <- which(abb.dataf$bedrooms > 3 & abb.dataf$type =='Apartment')
  ltra <- which(ltr.dataf$bedrooms > 3 & ltr.dataf$type =='Apartment')
  if(length(abba)>0) abb.dataf <- abb.dataf[-abba, ]
  if(length(ltra)>0) ltr.dataf <- ltr.dataf[-ltra, ]

@@

 ## Remove missing data
ltr.dataf <- ltr.dataf[!is.na(ltr.dataf$baths), ]
abb.dataf <- abb.dataf[!is.na(abb.dataf$baths), ]

@@

  ltbath <- ggplot(ltr.dataf, aes(x=baths)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:8, breaks=1:8)+
    xlab('\nBathrooms\n') +
    ylab('') + 
    ggtitle('Bathrooms in Long Term ltrals')

@@

  stbath <- ggplot(abb.dataf, aes(x=baths)) + 
    geom_bar(fill=abb.col[1], color=abb.col[1]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=0:8, breaks=0:8)+
    xlab('\nBathrooms\n') +
    ylab('') + 
    ggtitle('Bathrooms in Short Term ltrals')

@@

  # Long term data
  ltr.dataf <- ltr.dataf[ltr.dataf$baths < 4, ]

@@

In addition to looking at bedrooms and bathrooms as separate dimensions of properties, they can also be considered together.  In other words, it is very common in the industry (and when looking for a short term ltral) to specify searches by bed/bath combination for obvious reasons -- 1 bedroom with 3 baths is inefficient and 4 bedrooms with 1 bath is uncomfortable.  As a result we examine the combinations of the two.

@@

  ltr.dataf$bedbath <- paste0(ltr.dataf$bedrooms, '..', ltr.dataf$baths)
  abb.dataf$bedbath <- paste0(abb.dataf$bedrooms, '..', abb.dataf$baths)

@@

  ltb <- ggplot(ltr.dataf, aes(x=bedbath)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    xlab('\nBed/Bath Combo\n') +
    ylab('Number of Properties\n') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle('Long Term ltrals') + 
    facet_wrap(~type)

@@

  stb <- ggplot(abb.dataf, aes(x=bedbath)) +
    geom_bar(fill=abb.col[1], color=abb.col[1]) + 
    xlab('\nBed/Bath Combo\n') +
    ylab('Number of Properties\n') +
    ggtitle('Short Term ltrals') + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    facet_wrap(~type)

@@

  ltr.dataf <- ltr.dataf[ltr.dataf$bedbath %in% acc.bb, ]
  abb.dataf <- abb.dataf[abb.dataf$bedbath %in% acc.bb, ]

@@

### Rates and ltrs

Finally, we turn to filtering by the prices -- the nightly rates and weekly ltral values.  We start by removing observations missing price data.

@@

  ltr.dataf <- ltr.dataf[!is.na(ltr.dataf$event.price), ]
  abb.dataf <- abb.dataf[!is.na(abb.dataf$nightly.rate), ]

@@

  # ltr Plot
  ltr.plot <- ggplot(ltr.dataf, aes(x=event.price)) + 
    geom_density(fill=abb.col[3],alpha=.8) +
    ylab('') +
    xlab('Weekly ltr') + 
    ggtitle('Long Term ltrals') + 
    scale_x_continuous(limits=c(0, 1500)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank())

  # Nightly Rate plot
  rate.plot <- ggplot(abb.dataf, aes(x=nightly.rate)) + 
    geom_density(fill=abb.col[1], alpha=.8) + 
    ylab('') +
    xlab('Nightly Rate') +
    ggtitle('Short Term ltrals') + 
    scale_x_continuous(limits=c(0, 600)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank())
